---
title: "Data Munging, Visualization, and Feature Engineering"
author: "Scott McKean"
date: "2020-01-19"
output: html_document
---

This notebook provides an introduction to data munging in the tidyverse. Why
tidyverse you might ask? Personally, I find it a really nice way to learn the 
'grammar of graphics' and 'grammar of data' - probably the best intro in my 
opinion, which is why I love teaching it.

The agenda for this part of the bootcamp includes:

1. Loading data w/ readr
2. Dates and times w/ lubridate
3. Tidy data concepts w/ tidyr
4. Feature engineering w/ dplyr
5. Visualization w/ ggplot2

We will need the following libraries for this bootcamp

```{r}
library(lubridate)
library(tidyr)
library(dplyr)
library(ggplot2)
```

## Loading data w/ readr

The first thing we need to do with data is consistently load it. R has some 
basic functionality with `read.csv`, and data.table is the best for large
datasets, using `fread`, but here we are going to demo a load an inspect for
the demo data.

https://readr.tidyverse.org/
Check out the cheat sheet here, paying attention to the 'write' functions

My general formula for loading data:
1) Load it (does it fail?)
2) Run head()
3) Look at the dimensions
4) Check the column data types

```{r}
energy = readr::read_csv("./data/energy_dataset.csv")

#old school
head(energy)
dim(energy)
str(energy)
summary(energy)

#lazy school
colnames(energy)
energy
```
Here we are going to load the weather data, which is a much larger table.

```{r}
weather = readr::read_csv("./data/weather_features.csv")
summary(weather)

# how many cities are there?
weather %>% 
  pull(city_name) %>%
  factor() %>% 
  unique()
```
## Tidy data concepts w/ tidyr

There is a well defined concept of 'tidy data' which is well summarized in this paper: https://vita.had.co.nz/papers/tidy-data.pdf

To quote the paper, in tidy data:
1. Each variable forms a column.
2. Each observation forms a row.
3. Each type of observational unit forms a table.

The concept of 'wide' and 'long' tables is really important in data science,
and worth chatting about. In our data, we are fortunate that both `energy` and `weather` are long tables.

Long tables are pretty nice when doing summaries and graphs, but can be difficult
to use in machine learning models or summarize properly. For example, if you
wanted to facet a ggplot among several groups, you will need a long table.

We can pivot tables back and forth relatively easily in R using the tidyr package. Read more and check out the cheat sheet here: https://tidyr.tidyverse.org/

```{r}
# long frames are useful
ggplot(weather) +
  geom_histogram(aes(x=temp), bins=40) +
  facet_wrap(. ~ city_name, scales='free') +
  xlab('Temperature')
```
## Dates and times w/ lubridate

https://lubridate.tidyverse.org/

Working with dates is hard and constantly a struggle, especially when working
with anything excel has touched. I would really encourage everyone to look at and follow the ISO8601 format, and really understanding how dates are dealt with. 

Since the hackathon is really time-series data, we need to be sure that dates and times are encoded exactly the same.

```{r}
# filter to Barcelona
barcelona_weather = weather %>%
  filter(city_name == 'Barcelona')

ggplot(barcelona_weather) +
  geom_line(aes(x=dt_iso, y=temp))
```

## Visualization w/ ggplot2
ggplot2 (grammar of graphics plotting) is in my opinion the best user interface
for static plots. There are so many visualization tools that no single tool
will be the best, but ggplot2 often beats any other tool I use.

https://ggplot2.tidyverse.org/

I'd recommend looking at histograms, line plots, and the other wide range
of things ggplot can do. Let's do three 'experiments'

Does temperature correlate with pressure?
Does total load correlate with temperature?

```{r}
ggplot(weather) +
  geom_point(aes(x=pressure, y=temp, color=city_name))

# wait outliers?!
# can trial and error explore with xlim
ggplot(weather) +
  geom_point(aes(x=pressure, y=temp, color=city_name)) +
  xlim(900,1100)
```


```{r}
barcelona_temp = weather %>%
  filter(city_name == 'Barcelona') %>%
  select(time = dt_iso, temp)

total_load = energy %>%
  select(time, load='total load actual')

ts_merge = merge(barcelona_temp, total_load, on='time')

# doesn't look good!
ggplot(ts_merge) +
  geom_point(aes(x=temp, y=load)) 
```


