---
title: "Data Manipulation and Modeling"
author: "Scott McKean"
date: "2021-01-22"
output: html_document
---

This notebook builds on some of the visualization and R fundamentals and dives
into the world of data manipulation and modeling in R. The challenge you face
is difficult - time series are some of the hardest modeling problems 
because they are noisy, they have trends and seasonality, and they are autocorrelated.

I won't be able to teach time series analysis in an hour, but I can point you
to some techniques that might help you on your forecasting journey. R is an
amazing tool for time series modeling and investigation, you just need to know
where to look.

We are going to focus on two things: 
1) A grammar of data manipulation
2) Time series analysis

```{r}
# grammar of data
library(dplyr)
library(lubridate)
library(janitor)
library(ggplot2)
library(readr)

# time series
library(zoo)
library(forecast)
```

# Load Data

```{r}
energy = readr::read_csv("./data/energy_dataset.csv") %>%
  janitor::clean_names()
weather = readr::read_csv("./data/weather_features.csv") %>%
  janitor::clean_names()
```
# Grammar of Data

In order to ask the right questions, you need to be able to aggregate and 
disaggregate data. We use the dplyr package to generate some summary statistics
and look at our data.

The most important verbs you need to know:
- select
- filter
- mutate
- group_by
- summarize

Learn WAY more here:
https://dplyr.tidyverse.org/

```{r}
# Extract year, month, week, day of week 
energy <- energy %>%
  mutate(year = year(time)) %>%
  mutate(month = month(time)) %>%
  mutate(week = week(time)) %>%
  mutate(day = day(time)) %>%
  mutate(day_of_week = wday(time)) %>%
  mutate(date = lubridate::date(time))

# Now we can start plotting some interesting things
# yearly trend boxplot
ggplot(energy) +
  geom_boxplot(aes(y=total_load_actual, group=year))

# yearly trend timeseries w/ groups
day_of_week_summary = energy %>% 
  group_by(year, day_of_week) %>% 
  summarise_all(mean, na.rm=TRUE)

ggplot(day_of_week_summary) +
  geom_point(aes(x=day_of_week, y=total_load_actual, colour=year))

# be careful about sat/sun
# 2014 data is suspect (low observations)

# monthly trend
ggplot(energy) +
  geom_boxplot(aes(y=total_load_actual, group=month))
```
# Time Series

We are going to talk about three packages and approaches to forecasting

```{r}
# let's decompose actual electricity demand
# but start with daily observations
daily_demand = energy %>%
  group_by(date) %>%
  summarize_all(mean) %>%
  select(date, total_load_actual)
  
daily_demand_ts = as.ts(read.zoo(daily_demand))
daily_demand_ts

?decompose

# first problem - NAs!
# second problem - One 2014 observation!
daily_demand = energy %>%
  filter(date >= ymd('2015-01-01')) %>%
  group_by(date) %>%
  summarize_all(mean, na.rm=TRUE) %>%
  select(date, total_load_actual)

# make a 'time-series'. This is important
daily_ts = ts(
  daily_demand$total_load_actual, 
  start=c(2015,01,01), frequency=365.25
  )

# forecast is awesome - ggplot + time series
autoplot(daily_ts) +
  ggtitle('Actual Load') +
  ylab('kWh') +
  xlab('Date')
```

Let's talk about time series decomposition
https://buildmedia.readthedocs.org/media/pdf/a-little-book-of-r-for-time-series/latest/a-little-book-of-r-for-time-series.pdf

```{r}
daily_ts %>%
  stl(t.window=13, s.window="periodic", robust=TRUE) %>%
  autoplot()
```

Finally, let's talk about ARIMA models in three lines of code!
https://otexts.com/fpp2/


```{r}
# Fit an ARIMA model in three lines of code
fit = auto.arima(daily_ts)
checkresiduals(fit)
autoplot(forecast(fit))

# Try another style of model
daily_ts %>% 
  ets() %>% 
  forecast() %>% 
  autoplot()
```
